Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Review along section 5
Add comments
Pass over 5.1-5.4 done, looking really good.

\subsection{Porting Irregular GPU Algorithms to Cooperative Kernels}\label{sec:portingalgorithms}

\begin{table}
\small
\centering
\begin{tabular}{ l c c c }
app & resizing barriers & ckill - cfork & transmit vars\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 - 0 & 4\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 - 0 & 0\\
\rowcolor{Gray1}
bc & 3 / 6 & 0 - 0 & 3\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 - 0 & 0 \\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 - 0  & 4 \\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 - 0  & 4 \\
\rowcolor{Gray3}
octree & 0 / 0 & 1 - 1 & 0\\
\rowcolor{Gray3}
game & 0 / 0 & 1 - 1 & 0\\
\end{tabular} \\
\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} - Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} - Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} - work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\end{table}

\ADComment{Describe the algorithms that we ported.}

\HEComment{This is more or less already done in section
\ref{sec:programmingguidelines} , maybe no need to rewrite here?}

\ADComment{Give stats on how many variables had to be marked with
  $\transmit$ (i.e., occurred in the restoration context), how many
  barriers we changed to resizing, how many offer kill and offer fork
  calls we included.}  \TSComment{check out
  Table~\ref{tab:applications}. The barrier column shows how many of
  the original barriers were ported to resizing barriers. In all cases
  except bc we converted all of them. The only reason why bc is
  different is because barriers occur deeper in the call stack and our
  prototype cannot handle that. The algorithm itself is compatable
  (i.e. changing these barriers is as easy as any other, there is
  nothing special about them.)}

\ADComment{Make sure we address the point that placing the primitives was pretty easy.}

\subsection{The GPUs we Tested}

\ADComment{Describe the platforms.  Explain why we could not try other
platforms.}

Our implementation approach requires two optional features of OpenCL
$2.0$, namely SVM fine-grained buffers and SVM atomics, which are
currently not widely supported. \nvidia chips are ruled out since
\nvidia does not support OpenCL $2.0$. Intel drivers meet our
requirements on Windows, but not on Linux, although we tried the
upstream driver that requires a patched Linux kernel. AMD support SVM
atomics only their Kaveri family of chips and only on Linux, requiring
to compile the latest driver. Among the GPUs available to us, only 4
turned out to provide our requirements: Intel HD 520, HD 5500 and IRIS
6100, and AMD Radeon R7.

\HEComment{To decide: do we mention carrot AMD, in which terms?}

@@

A cooperative kernel can yield workgroups to let a non-cooperative,
transient one execute. We size non-cooperative kernels to represent
approximatively the workload required by a system to maintain smooth
screen animations.

\paragraph{Experimental setup} We repetitively launch a kernel with a
controlled workload and detect from which workload size the screen
animation starts to jitter. Given the kernel workload and its
scheduling frequency, we can deduce what is the workload and frequency
needed by the system to maintain a smooth animation. We use this
approach to measure the system workloads needed to maintain smooth
animations of three types: (\emph{light}) emphasis of desktop icons
under the mouse pointer, (\emph{medium}) window dragging over the
desktop, and (\emph{heavy}) WebGL animations in a browser.

\begin{table}
\small
\centering
\begin{tabular}{ l c c }
Workload & Duration (ms) & Period (ms)\\
\hline
light & 3 & 70\\
medium & 3 & 40\\
heavy & 10 & 40\\
\end{tabular}
\caption{Caracteristics of non-cooperative workloads}
\label{tab:noncooperativeworkload}
\end{table}

@@

\subsection{The Overhead of Moving to Cooperative Kernels}

@@

\hline
\end{tabular}\\
{\footnotesize
$^{\ast}$color ecology, $^{\dagger}$bc 1k\_128k, $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
}
\caption{Overhead of cooperative kernels: execution time slowdown}
\label{tab:overhead}
\end{table}

@@

\paragraph{Experimental setup}
Cooperative scheduling primitives entails an overhead even if no
killing, forking or resizing happen. We measure this overhead via the
execution time slowdown between the original kernel and the
cooperative one, forcing the scheduler to never modify the number of
active workgroups. Since the cooperative kernel contains the
non-cooperative kernel code anyway, its occupancy bound may be less
than the original kernel; we make sure to use this occupancy bound as
the number of workgroups for the original kernel in order to compare
relevant execution times. \HEComment{Do we need to mention the
  possible occupancy reduction as an overhead?}

\paragraph{Results}
\mytab~\ref{tab:overhead} lists both geometric mean and maximum value
of measured overheads. The slowdown is below 2 even in the worst case,
and closer to 1 in general. We consider these results as encouraging,
especially since our prototype performance is likely a far upper bound
of what a scheduler implemented directly in the runtime could achieve.

@@

voc: blocking and non-blocking ?

durable / transient

@@

Its performance corresponds

@@

Cooperative scheduling can improve performances

>>>>

We discuss our experience porting irregular algorithms to cooperative kernels (\mysec\ref{sec:portingalgorithms}) and discuss the GPUs on which we evaluate these applications (\mysec\ref{sec:gpus}).  For these GPUs, we report on experiments to determine non-cooperative workloads that model the requirements of various graphics rendering tasks (\mysec\ref{sec:sizingnoncoop}).  We then examine the overhead associated with moving to cooperative kernels when multitasking is \emph{not} required (\mysec\ref{sec:overhead}), as well as the responsiveness and throughput observed when a cooperative kernel is multi-tasked with non-cooperative workloads (\mysec\ref{sec:responsiveness}).

\subsection{Porting Irregular GPU Algorithms}\label{sec:portingalgorithms}

\begin{table}
\small
\centering
\begin{tabular}{ l c c c c }
App. & resizing barriers & kill & fork & transmit vars\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0\\
\rowcolor{Gray1}
bc & 3 / 6 & 0 & 0 & 3\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0 \\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4 \\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4 \\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0\\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0\\
\end{tabular} \\
\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\end{table}

\ADComment{It might be good to add a LOC column to \mytab\ref{tab:applications} to indicate how large the original versions of these applicaitons were.}

\ADComment{I thought we had some examples where we actually wanted a non-resizing barrier; we say this earlier in the paper so we need to smooth.}

\ADComment{Missing: a discussion of what we did not port.}

\ADComment{We need to say something about the number of data sets
  associated with each application, so that we know the total number
  of application/data-set pairs.}

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...} GPU application suites,
using global barriers.  We indicate how many of the original number of barriers are changed to resizing barriers, and how many variables need to
be transmitted.  In all cases all barriers were converted, except for bc
which contains barriers deeper in the call stack, a case not supported
by our prototype. Still, these barriers could in principle also be
converted.  The remaining 2 algorithms are workstealing applications:
each required the addition of $\offerfork$ and $\offerkill$ at the start
of the main loop, and no variables needed to be transmitted. For all
these examples the porting to cooperative kernels was similarly straightforward to the porting illustrated in \mysec\ref{sec:programmingguidelines} for our running examples, and did not require reshaping of the original algorithm.

\subsection{The GPUs we Tested}\label{sec:gpus}

Our prototype implementation (\mysec\ref{sec:implementation}) requires two optional features of OpenCL
2.0: SVM fine-grained buffers and SVM atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at present, since
these vendors do not yet support OpenCL $2.0$, equivalent functionality is not provided by \nvidia's CUDA. Intel drivers meet our
requirements on Windows, but not on Linux (we confirmed this by trying the
upstream driver that requires a patched Linux kernel). AMD support SVM
atomics only their Kaveri family of chips and only on Linux, requiring
to compile the latest driver.

\ADComment{Check:} Among the GPUs available to us, four met our
requirements: Intel HD 520, HD 5500 and Iris 6100, and AMD Radeon R7.
However, AMD's Linux drivers for OpenCL suffer from a known defect
whereby long-running kernels lead to defunct processes that the OS
cannot kill.  We observed this on our AMD platform with the latest
driver, and simultaneously obtained alarming results on this platform
when using SVM atomics: modifying a kernel to include a simple
host/device handshake through SVM slowed down kernel execution by an
order of magnitude in some cases.  This behaviour, combined with the
defunt processes issue, lead us to suspect that the support we require
for cooperative kernels is not yet mature enough within AMD's drivers,
which we plan to discuss with engineers at AMD.

We thus present experimental results for the Intel GPUs, whose characteristics are summarised in \mytab\ref{tab:chipstested}.

\begin{table}
\small
\centering
\begin{tabular}{ l l l l }
Chip & HD520 & HD5500 & Iris\\
\hline
\# compute units  & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
Driver   & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
Host CPU & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
\end{tabular}
\caption{The Intel GPU platforms used in our evaluation}
\label{tab:chipstested}
\end{table}

@@

\ADComment{Refer back to nonfunctional requirements section, \mysec\ref{sec:nonfunctional}.}

Enabling rendering of smooth graphics in parallel with execution of irregular algorithms is an important use case for our approach.  Because our prototype implementation is based on a mega-kernel that takes over the entire GPU (see \mysec\ref{sec:implementation}), we cannot assess this use case directly.

As a proxy for genuine graphics rendering, we devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
\ADComment{I think this is good but could be clearer and it would be helpful to discuss.}
We repetitively launch a kernel with a
controlled workload and detect from which workload size the screen
animation starts to jitter. \ADComment{Be clear that this is manual.} Given the kernel workload and its
scheduling frequency, we can deduce what is the workload and frequency
needed by the system to maintain a smooth animation. We use this
approach to measure the system workloads needed to maintain smooth
animations of three types: (\emph{light}) emphasis of desktop icons
under the mouse pointer, (\emph{medium}) window dragging over the
desktop, and (\emph{heavy}) WebGL animations in a browser.

\begin{table}
\small
\centering
\begin{tabular}{ l r r }
Workload & Duration (ms) & Period (ms)\\
\hline
light & 3 & 70\\
medium & 3 & 40\\
heavy & 10 & 40\\
\end{tabular}
\caption{\ADComment{Clearer caption?} Characteristics of non-cooperative kernels workload}
\label{tab:noncooperativeworkload}
\end{table}

@@

\subsection{The Overhead of Moving to Cooperative Kernels}\label{sec:overhead}

@@

\hline
\end{tabular}\\
{\footnotesize
$^{\ast}$color ecology, $^{\dagger}$bc 1k\_128k, $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
}
\caption{Execution time slowdown associated with cooperative scheduling mode without multi-tasking}
\label{tab:overhead}
\end{table}

@@

\paragraph{Experimental setup}
Invoking the cooperative scheduling primitives incurs some overhead
even if no killing, forking or resizing actually occurs, because the cooperative kernel still needs to interact with the scheduler to determine this.
We assess this overhead by measuring the
 slowdown in execution time between the original and cooperative versions of a kernel, forcing the scheduler to never modify the number of
active workgroups in the cooperative case.

Recall that our mega kernel-based implementation merges the code for a
cooperative kernel with that of the non-cooperative task to be
co-scheduled.  This blow-up in code size can have a negative impact on
compiliation, e.g.\ due to higher register pressure, reducing the
occupancy for the merged kernel---the maximum number of workgroups
that the GPU driver will schedule.  This is an artifact of our
proof-of-concept implementation strategy, and would not be a problem
if our cooperative scheduling approach was implemented inside the GPU
driver.  We thus launch both the original and cooperative versions of
a kernel with the reduced occupancy bound in order to meaningfully
compare execution times.

\ADComment{How many times did we run each benchmark?}

\paragraph{Results}
For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
maximum slowdown across all applications and data sets.  For the
maximum slowdowns, we indicate which application and data set was
responsible.  The slowdown is below 2 even in the worst case, and
closer to 1 on average. We consider these results as encouraging,
especially since the performance of our prototype could clearly be
improved upon if cooperative scheduling was implemented directly in
the GPU driver.

\ADComment{Can we remark a bit more on the outliers?}

@@

\HEComment{Below, just random notes so far}

@@

@@
